Comparing fluorometric methods (in-vivo vs extracted) for cyanoHAB monitoring in six Rhode Island ponds

### 2023-12-04 manuscript/manuscript.qmd
Jeff edits on paper.  Ready for Betty review?!?

---
title: "Comparing fluorometric methods (in-vivo vs extracted) for cyanoHAB monitoring in six Rhode Island ponds"
author: 
  - name: Stephen D. Shivers
    orcid: 0000-0003-0096-7433
    email: shivers.stephen@epa.gov
    corresponding: true
    affiliation:
      - id: epa
        name: US Environmental Protection Agency, Atlantic Coastal Environmental Sciences Division
        city: Narragansett
        state: RI
        zip: 02882
  - name: Jeffrey W. Hollister
    orcid: 0000-0002-9254-9740
    affiliation:
      - ref: epa
  - name: Sophie Fournier
    affiliation:
      - id: orise
        name: Oak Ridge Institute for Science and Education
        city: Narragansett
        state: RI
        zip: 02882
  - name: Jakob Stankoski
    affiliation:
      - id: orau
        name: Oak Ridge Associated Universities
        city: Narragansett
        state: RI
        zip: 02882
  - name: Betty J. Kreakie
    orcid: 0000-0002-4982-7610
    affiliation:
      - ref: epa
abstract: "Harmful algal blooms caused by cyanobacteria (cyanoHABs) are detrimental to human and environmental health and can be difficult to monitor without specialized training and equipment.  A variety of instruments have been developed to measure cyanoHAB indicators (i.e. chlorophyll *a* or phycocyanin) that do not require an extraction process.  We compared measurements from five in-vivo fluorometers (Turner Trilogy in-vivo module, Turner Fluorosense, Turner Cyanofluor, bbe AlgaeTorch, and bbe Phycoprobe) to results from solvent-based extractions for chlorophyll *a* and phycocyanin at six different waterbodies in Rhode Island.  We found a strong relationship between extracted phycocyanin and in-vivo fluorometers (r2 ranging from 0.78-0.96).  We found less consistency between in-vivo measurements of chlorophyll *a* and the extracted results (r2 between 0.34 and 0.82).  Some variability in the chlorophyll *a* results can be explained by differences in the phytoplankton community at the different sampling locations.  Phycocyanin results from in-vivo fluorometry were also strongly correlated with cell counts, which implies that phycocyanin measurements from these instruments can be a good proxy for cell counts.  Many federal, state, and local entities use cell counts of cyanobacteria to determine when to issue health or contact advisories for waterbodies.  Producing accurate cell counts requires highly specialized training/equipment, processing time, and counts can vary greatly between technicians.  The results from this study encourage further research using in-vivo fluorometry and discussion to move away from cell counts and towards fluorometric based measurements when making decisions regarding public health."
format: 
    docx:
      reference-doc: manuscript_template.docx
bibliography: manuscript.bib
---

@@

Field instruments may be placed directly in a waterbody or require water to be poured into a cuvette and then placed into a handheld fluorometer. Differences in measurement methods represent a potential source of error when comparing instruments. Additionally, instruments may report valued in rfus (raw or relative fluorescence units) or µg/L (concentration) and comparing these values introduces an additional source of error during comparison. In vivo fluorescence can be affected by photochemical quenching causing an underestimation of rfus as photosystems become saturated with light. Estimation problems can also result from differing community composition, particularly if colonial cyanobacteria are present as the interior of large colonies will not fluoresce as readily as the exterior. Finally, the physical properties of the sample water can greatly affect fluorescence readings based on turbidity levels and the presence of CDOM. Some instruments offer a yellow substance correction to help account for CDOM while others do not. How the different instruments handle these various potential problems can affect their accuracy and precision, and the ability to compare across different instrument types. The goal of this paper is to compare different handhelds or in situ fluorometers with respect to a common benchmark (extracted chlorophyll and phycocyanin) across a variety of waterbodies.

@@

## Summary of Data

<!-- Jeff -->

Need summary of sites and dataFig 1. Map of Field Sites

@@

Extracted chlorophyll measurements ranged from 3.2 ug/L (Indian Lake ) to 81.3 ug/L (Mashapaug Pond). Chlorophyll measurements for all instruments followed the same general pattern as the extracted chlorophyll values (@fig-chlascatter).  The Algaetorch and Phycoprobe measurements exhibited a strong relationship with extracted chlorophyll concentrations with R2 values of 0.82 and 0.81, respectively.  The Cyanofluor (R2=0.35) and Trilogy in vivo (R2=0.34) measurements increased as extracted chlorophyll concentrations increased, but the relationship was not as strong compared to the other instruments.

@@

Extracted phycocyanin measurements ranged from 1.3 ug/L (Indian Lake) to 75.0 ug/L (Mashapaug Pond). The Algaetorch (r2=0.86) and Phycoprobe (r2=0.84) were strongly related to extracted phycocyanin concentrations but were not directly comparable as both measurements were of cyanobacteria as chlorophyll instead of a direct measure of phycocyanin (@fig-phycoscatter). Phycocyanin measurements for the Fluorosense (r2=0.96) and the Cyanofluor (r2=0.78) were also strongly related with extracted phycocyanin.

@@

Extracted chlorophyll *a* was the most strongly correlated with cyanobacterial cell counts (r2=0.92) followed by the algaetorch (r2=0.78) and the Phycoprobe (r2=0.59). The Cyanofluor (r2=0.12) and Trilogy in vivo module (0.11) were weakly correlated (@fig-chlcount). Phycocyanin measurements from all instruments were strongly correlated with cell counts (r2 between 0.9 and 0.97)(@fig-phycocount).

@@

The Algaetorch and Phycoprobe yielded similar measurements and also overestimated phycocyanin concentrations but the output from these instruments is not a direct measure of phycocyanin; instead, these instruments measure the percentage of the chlorophyll that is derived from cyanobacteria. The Cyanofluor output was in rfus and was not directly comparable to extracted phycocyanin without creating a standard curve for the instrument using known phycocyanin standard concentrations, which was beyond the scope of this study. A study by Thomson-Laing et al. did convert rfus to ug/L and found strong correlation between cyanobacteria biovolume and phycocyanin concentration for eleven cyanobacteria cultures (Thomson-Laing et al. 2020). They also found stronger relationships for environmental samples in lakes that were cyanobacteria dominated compared to more diverse lakes.

All of these instruments could be used to effectively monitor for the presence of cyanobacteria by measuring more direct parameters of cyanobacteria than chlorophyll a. All of the fluorometers exhibited a positive relationship with chlorophyll, but the Algaetorch and Phycoprobe had a much stronger relationship. Reduced correlation by the Cyanofluor and Trilogy in vivo module was driven by the underestimation of chlorophyll in Mashapaug Pond and possible overestimation in Warwick and Curran. Phycocyanin concentrations and total biovolume of cyanobacteria were the highest in Mashapaug and it is likely that these high concentrations were underestimating the in vivo concentrations of chlorophyll (Zamyadi et al . 2012). Therefore, the Algaetorch and Phycoprobe are more successful in properly estimating chlorophyll across a range of different waterbodies, but they are also the most expensive instruments. A study by Silva et al. found that the Fluoroprobe (a similar instrument to the Phycoprobe) was correlated with cell counts at chlorophyll *a* concentrations below 100 ug/L but not above, but the present study did not have study ponds that exceed 100 ug/L (Silva et al. 2016).

@@

# Summary

1.) In general, in situ phycocyanin better match to extracted than in situ chlorophyll is to extracted. The cholorophylll relationship is driven by phytoplankton community make up

2.) Extracted chlorophyll and phycocyanin, are great match to cell counts. Suggests pigments are a good proxy for cell counts and could be used in place of cell counts.

3.) A discussion with water quality managers about using extracted or in situ phycocyanin as a replacement for cell counts might be warranted.

Don't read anything into absolute values (i.e. chlorophyll from one sensor not equal to chlorophyll from another sensor) across sensors, but general patterns are still very useful. Need to know info about specific sites to be useful.

Phycocyanin very good proxy for cell counts. Suggest replacing cell counts for waterbody listing decisions with phycocyanin. Would likely need additional work to furhter explore relationship, build predictive models of cell coutns, etc.

@@

And special thanks to Gracie, Anastasia, and Lilac for emotional support during challenging analysis times.

>>>>

---
title: "Comparing fluorometric methods (in-vivo vs extracted) for cyanoHAB monitoring in six Rhode Island ponds"
author: 
  - name: Stephen D. Shivers
    orcid: 0000-0003-0096-7433
    email: shivers.stephen@epa.gov
    corresponding: true
    affiliation:
      - id: epa
        name: US Environmental Protection Agency, Atlantic Coastal Environmental Sciences Division
        city: Narragansett
        state: RI
        zip: 02882
  - name: Jeffrey W. Hollister
    orcid: 0000-0002-9254-9740
    affiliation:
      - ref: epa
  - name: Sophie Fournier
    affiliation:
      - id: orise
        name: Oak Ridge Institute for Science and Education
        city: Narragansett
        state: RI
        zip: 02882
  - name: Jakob Stankoski
    affiliation:
      - id: orau
        name: Oak Ridge Associated Universities
        city: Narragansett
        state: RI
        zip: 02882
  - name: Betty J. Kreakie
    orcid: 0000-0002-4982-7610
    affiliation:
      - ref: epa
abstract: "Harmful algal blooms caused by cyanobacteria (cyanoHABs) are detrimental to human and environmental health and can be difficult to monitor without specialized training and equipment.  A variety of instruments have been developed to measure cyanoHAB indicators (i.e. chlorophyll *a* or phycocyanin) that do not require an extraction process.  We compared measurements from five in-vivo fluorometers (Turner Trilogy in-vivo module, Turner Fluorosense, Turner Cyanofluor, bbe AlgaeTorch, and bbe Phycoprobe) to results from solvent-based extractions for chlorophyll *a* and phycocyanin at six different waterbodies in Rhode Island.  We found a strong relationship between extracted phycocyanin and in-vivo fluorometers (R^2^ ranging from 0.78-0.96).  We found less consistency between in-vivo measurements of chlorophyll *a* and the extracted results (R^2^ between 0.34 and 0.82).  Some variability in the chlorophyll *a* results can be explained by differences in the phytoplankton community at the different sampling locations.  Phycocyanin results from in-vivo fluorometry were also strongly correlated with cell counts, which implies that phycocyanin measurements from these instruments can be a good proxy for cell counts.  Many federal, state, and local entities use cell counts of cyanobacteria to determine when to issue health or contact advisories for waterbodies.  Producing accurate cell counts requires highly specialized training/equipment, processing time, and counts can vary greatly between technicians.  The results from this study encourage further research using in-vivo fluorometry and discussion to move away from cell counts and towards fluorometric based measurements when making decisions regarding public health."
format: 
    docx:
      reference-doc: manuscript_template.docx
bibliography: manuscript.bib
---

@@

Field instruments may be placed directly in a waterbody or require water to be poured into a cuvette and then placed into a handheld fluorometer. Differences in measurement methods represent a potential source of error when comparing instruments. Additionally, instruments may report valued in RFUs (raw or relative fluorescence units) or µg/L (concentration) and comparing these values introduces an additional source of error during comparison. In vivo fluorescence can be affected by photochemical quenching causing an underestimation of RFUs as photosystems become saturated with light. Estimation problems can also result from differing community composition, particularly if colonial cyanobacteria are present as the interior of large colonies will not fluoresce as readily as the exterior. Finally, the physical properties of the sample water can greatly affect fluorescence readings based on turbidity levels and the presence of CDOM. Some instruments offer a yellow substance correction to help account for CDOM while others do not. How the different instruments handle these various potential problems can affect their accuracy and precision, and the ability to compare across different instrument types. The goal of this paper is to compare different handhelds or in situ fluorometers with respect to a common benchmark (extracted chlorophyll and phycocyanin) across a variety of waterbodies.

@@

@@

Extracted chlorophyll measurements ranged from 3.2 ug/L (Indian Lake ) to 81.3 ug/L (Mashapaug Pond). Chlorophyll measurements for all instruments followed the same general pattern as the extracted chlorophyll values (@fig-chlascatter).  The Algaetorch and Phycoprobe measurements exhibited a strong relationship with extracted chlorophyll concentrations with R^2^ values of 0.82 and 0.81, respectively.  The Cyanofluor (R^2^=0.35) and Trilogy in vivo (R^2^=0.34) measurements increased as extracted chlorophyll concentrations increased, but the relationship was not as strong compared to the other instruments.

@@

Extracted phycocyanin measurements ranged from 1.3 ug/L (Indian Lake) to 75.0 ug/L (Mashapaug Pond). The Algaetorch (R^2^=0.86) and Phycoprobe (R^2^=0.84) were strongly related to extracted phycocyanin concentrations but were not directly comparable as both measurements were of cyanobacteria as chlorophyll instead of a direct measure of phycocyanin (@fig-phycoscatter). Phycocyanin measurements for the Fluorosense (R^2^=0.96) and the Cyanofluor (R^2^=0.78) were also strongly related with extracted phycocyanin.

@@

Extracted chlorophyll *a* was the most strongly correlated with cyanobacterial cell counts (R^2^=0.92) followed by the algaetorch (R^2^=0.78) and the Phycoprobe (R^2^=0.59). The Cyanofluor (R^2^=0.12) and Trilogy in vivo module (0.11) were weakly correlated (@fig-chlcount). Phycocyanin measurements from all instruments were strongly correlated with cell counts (R^2^ between 0.9 and 0.97)(@fig-phycocount).

@@

The Algaetorch and Phycoprobe yielded similar measurements and also overestimated phycocyanin concentrations but the output from these instruments is not a direct measure of phycocyanin; instead, these instruments measure the percentage of the chlorophyll that is derived from cyanobacteria. The Cyanofluor output was in RFUs and was not directly comparable to extracted phycocyanin without creating a standard curve for the instrument using known phycocyanin standard concentrations, which was beyond the scope of this study. A study by Thomson-Laing et al. did convert RFUs to ug/L and found strong correlation between cyanobacteria biovolume and phycocyanin concentration for eleven cyanobacteria cultures (Thomson-Laing et al. 2020). They also found stronger relationships for environmental samples in lakes that were cyanobacteria dominated compared to more diverse lakes.

All of these instruments could be used to effectively monitor for the presence of cyanobacteria by measuring more direct parameters of cyanobacteria than chlorophyll *a*. All of the fluorometers exhibited a positive relationship with chlorophyll, but the Algaetorch and Phycoprobe had a much stronger relationship. Reduced correlation by the Cyanofluor and Trilogy in vivo module was driven by the underestimation of chlorophyll in Mashapaug Pond and possible overestimation in Warwick and Curran. Phycocyanin concentrations and total biovolume of cyanobacteria were the highest in Mashapaug and it is likely that these high concentrations were underestimating the in vivo concentrations of chlorophyll (Zamyadi et al . 2012). Therefore, the Algaetorch and Phycoprobe are more successful in properly estimating chlorophyll across a range of different waterbodies, but they are also the most expensive instruments. A study by Silva et al. found that the Fluoroprobe (a similar instrument to the Phycoprobe) was correlated with cell counts at chlorophyll *a* concentrations below 100 ug/L but not above, but the present study did not have study ponds that exceed 100 ug/L (Silva et al. 2016).

@@

@@

We are grateful to EPA Reviewer 1, EPA Reviewer 2, EPA Reviewer 3, ACESD QAM, ACESD Supervisor, and Tim Gleason for constructive reviews of this paper.  The views expressed in this article are those of the authors and do not necessarily represent the views or policies of the U.S. Environmental Protection Agency. Any mention of trade names, products, or services does not imply an endorsement by the U.S. Government or the U.S. Environmental Protection Agency. The EPA does not endorse any commercial products, services, or enterprises. This contribution is identified by the tracking number ORD-XXXXXX of the Atlantic Coastal Environmental Sciences Division, Office of Research and Development, Center for Environmental Measurement and Modeling, US Environmental Protection Agency. Lastly, special thanks to Gracie, Anastasia, and Lilac for cheer leading and support during challenging analysis times.
